Time invariant analysis of epidemics with \pkg{EpiCompare}

### 2021-01-26 jss/EpiCompare-paper/EpiCompare-paper.tex
rcpp speed up, clean up of arrangeGrob + grid.arrange, and additional comment in readme
section 3
epicompare section 2 rewrite + update of pipeline2 image

\Abstract{
We present \pkg{EpiCompare}, an \proglang{R} package that suppliments
and enhance current infectious disease modeling analysis pipelines as
well as to encourage comparisons across these pipelines. A major
contribution of this work is the set of novel \textit{time-invariate}
tools for model and epidemic comparisons - including time-invariate
prediction bands. \pkg{EpiCompare} encorporates \proglang{R}'s
\textit{tidy} coding style to aid it rapid and easy use. This paper
provides an overview of both the tools in and intuition behind
\pkg{EpiCompare} and a thorough demonstrating of the tools through a
detailed example of a full data analysis pipeline.
}

@@

\usepackage{amsmath} \usepackage{amsthm}

@@

@@

\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way. Epidemics, despite by definition being a
process that evolves over time, often need to be compared in a way not
constrained to initial times or time scales to understand the processes
at play. Additionally, many tools designed to examine the quantity of
the population in each epidemic state (for example: quantity of
Susceptible vs Infectious vs Recovered individuals) don't always as
intelligently capture the natural connections between the proportion of
individuals in these states.
\textcolor{orange}{I don't quite understand this.  Is this because of underreporting?  If so I'm not sure how EpiCompare helps us here.}
Tools in \pkg{EpiCompare} give the user the ability to extend their
toolkit to evaluate epidemics within a time-invariant lens. The goal of
\pkg{EpiCompare} is not to supplant existing infectious disease modeling
tools and software but, rather, is a concerted effort to create standard
and fair comparisons among models developed for disease outbreaks and
outbreak data.

This paper is broken up into the following sections, section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a thorough demonstrating of the tools
through a detailed example of a full data analysis pipeline.

@@

Epidemics can be difficult to compare to one another due to differences
in diseases, locations, or population behaviors, or times. In
\pkg{EpiCompare}, we emphasize comparisons between epidemics adjusting
for the last component, time. Time-invariant analysis is beneficial
because by adjusting for the unit of infection rate, we can focus on the
``lifetime'' of an epidemic, a view that is concerned more with the
number of lives affected than as opposed to any specific time
constraint. Time-invariant analysis is also beneficial when there are
gaps in time between occurrences of outbreaks of a similar nature in
different geographic regions. Finally, time-invariant analysis is
beneficial because many studies talk about a period of ``exponential
growth'\,' of the number of outbreaks
\citep{chowell2007,wallinga2007generation,forsberg2008}.

Time-invariant analysis, as it appears in \pkg{EpiCompare} solves the
above issues by observing functionals which attempt to capture the
general shape of the epidemic with respect to the proportion of the
population in each epidemic state. This allows us to compare scenarios
as different as, for instance as a decades-long outbreak HIV in the US
compared to a 10 day outbreak of norovirus on a cruise ship. Moreover,
this tool avoids the need for choosing a `beginning' \(t_0\) or `end'
\(t_F\) of an epidemic, choices \cite{gallagher2020} show that can
heavily influence, for example, estimates of peak infection height or
the reproduction number \(R_0\).

\subsection[r0]{\(R_0\) and time-invariant analysis}\label{r0}

By definition, \(R_0\) is the number of expected secondary infections
when a primary infection is introduced to a susceptible population.
\(R_0\) is also, maybe, the most famous \textit{time-invariant}
numerical summary of an epidemic, which allows epidemics to be compared
to one another in both different time and geographic scales. For
example, \(R_0\) for Covid-19 is estimated to be between 2-3, seasonal
influenza between 1.2-2, and modern measles outbreaks as large as 12
\citep{midas2020,biggerstaff2014,namee2018}.

Estimators for \(R_0\) are dependent on the epidemic modeling framework,
which consists of which states an individual can occupy
(e.g.~susceptible, infectious, recovered) and a description of how
individuals move from one state to the next over time (see
\cite{hethcote1994}). A common epidemic modeling framework is the SIR
model, originally introduced by \citet{kermack1927}. Transitions from
one state to the next are defined by a series of ordinary differential
equations, where \(N\) is the (fixed) total number individuals,
\(\beta\) is the rate of infection, and \(\gamma\) is rate of recovery,
\begin{align}\label{eq:sir-ode}
      S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
      I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
      R^\prime(t) &= \gamma I(t) \nonumber.
  \end{align} From this,
\(\hat{R}_0 = \frac{\hat{\beta}}{\hat{\gamma}}\), the ratio of the
estimated infection rate compared to the estimated recovery rate.

With regards to traditional epidemic \(state\) vs.~\(time\) plots,
\(R_0\) is difficult to visualize, especially with respect from one
epidemic to another. For example, consider the scenarios where the first
epidemic is generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_1 = 0.3\) and \(\gamma_1 = 0.15\), and the second epidemic is
generated from a SIR model with \((S(0) = 990, I(0) = 10)\),
\(\beta_2 = 0.24\) and \(\gamma_1 = 0.12\) over the first 15 days. Both
epidemics have the same value of
\(R_0 = \beta_1/ \gamma_1 = \beta_2 / \gamma_2 = 2\). The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:different-scales-standard}. At a glance, we may assume that
Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
occurs more quickly than in Model 2. On the other hand, we may think
Model 2 has a larger \(R_0\) because we may think the number of
infections in that model has not yet peaked at time 15.

@@

However, when we plot the trajectories as a single curve using the
ternary plot in the time-invariant view, we immediately see a different
story. In this time-invariant view in Fig.
\ref{fig:different-scales-tern}, the points seem to overlap and form the
same trajectory. Now it seems to be that Model 2 is following the same
trajectory as Model 1 but is not as far along in the infection process.
We can see there is something fundamentally linking these two different
epidemics, and this fundamental link turns out to be \(R_0\).

More formally, let two Kermack and McKendrick SIR models(see
Eq.\eqref{eq:sir-ode} ) be denoted \((S_1(t), I_1(t), R_1(t))\) and
\((S_2(t), I_2(t), R_2(t))\), respectively, for \(t > 0\). Assume both
models have initial values \((S(0), I(0), R(0))\). Let
\(R_0 = \frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). Equivalently,
\(\beta_2 = a \beta_1\) if and only if \(\gamma_2 = a \gamma_1\) for
some \(a > 0\).

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model is also a point on the epidemic path of the second SIR
model. Taking the sample simulations from Fig.
\ref{fig:different-scales-standard}, Fig.
\ref{fig:different-scales-tern} presents these two models in a ternary
plot. This means with our time-invariant ternary plot, that at a glance,
we can tell if two epidemics have different values of \(R_0\).

{\centering \includegraphics{Figs/unnamed-chunk-3-1}

\caption{\label{fig:different-scales-tern}Example of two epidemics with different $\beta$ and $\gamma$ paremeters but the same initial reproduction number $R_0$ = 2.  Both plots are generated from models with $N= 1000$ individuals with $S(0) = 990$ and $I(0) = 10$.  These are plotted in the time-invariant view, where we can see the number of susceptible, infectious, and recovered.}\label{fig:unnamed-chunk-3}
\end{figure}
\end{CodeChunk}

\hypertarget{beyond-the-kermack-and-mckendrick-sir-models}{%
\subsection{Beyond the Kermack and McKendrick SIR
models}\label{beyond-the-kermack-and-mckendrick-sir-models}}

Although the result of Theorem \ref{thm:sir-scale} allows for easy
visual comparison of \(R_0\) in Kermack SIR models, it does require
stringent assumptions of homogeneity of behavior in populations. The use
of visualizing epidemics in a time-invariant lens via ternary plots
extends beyond those of models that follow the ODEs in the
Kermack-McKendrick equations. Any model with S, I, and R states can be
visualized with ternary plots, regardless of birth and death dynamics
and regardless of homogeneity of individual behavior. We can use ternary
plots to compare the spread of a disease for groups within a population
without time as a confounding factor. We show an example of this in a
later section.

Moreover, time-invariant analysis is also useful for epidemic models
with more than three epidemic states. The constraints in three
dimensions that are met with the SIR model (there are three epidemic
states and for each time step, the sum of total number of individuals in
each state is the population size at that time) represents a space of 3d
simplices, and the ternary plot specifically represents these simplices,
after scaling the values in each state as proportions of the total
population at that time (ternary plots are known as a 3d unit simplex
due to it's scaling). This same scaling for larger models (i.e.~with
more states) can be done onto different simplexes. In this package we
present tools to help compare models (mostly through simulations).
\pkg{EpiCompare}'s tools permits comparisons of \(d\)-dimensional
objects by projecting them into (\(d-1\))-dimensional space.

Even though higher dimensional models may not be able to visualized
simply, we provide a number of tools to aid in the comparison of models
and epidemics. The first of which uses multiple simulations under
specific model parameters to assess the \textcolor{orange}{variability?}
bairabilty of the model fit. In \pkg{EpiCompare}, we provide ways to
create prediction regions for a `true' epidemic under a fully specified
model. These regions require representing multi-dimensional structures
for functions to completely contain epidemics. \pkg{EpiCompare} treats
these simulations and epidemics as \textit{filamental} objects --
\textcolor{orange}{need a description for filamental, Ben feel free to change}
where not only the outbreak trajectory points are important but also
their form which can be considered a function.
\textcolor{orange}{could be useful to have a picture/example here? like  three curves and 2 metrics one where closer under a pointwise distance and the other closer in filamental distance? I'll sketch up an idea unless you have something in your proposal already}

We extend off of papers like \citet{Dalmasso2019a} to create these
bands. These high dimensional bands allow the user to assess if the true
epidemic is within the band (thereby assessing the model's
representation of the epidemic). With these bands, we can compare
\textit{models} to \textit{models} by assessing the distance between the
two bands of model representation. We recommend using the Hausdorff
distance to compare two bands to one another, as this distance captures
how much bigger the sets would have to expand to cover each other. The
Hausdorff distance is defined mathematically as \[
d_\text{Hausdorff}(S_1, S_2) = \max \left\{ \sup_{x \in S_1} \inf_{y \in S_2} d(x,y),\; \sup_{y \in S_2} \inf_{x \in S_1} d(x,y)\right\}\;.
\]

>>>>

\Abstract{
We present \pkg{EpiCompare}, an \proglang{R} package that suppliments
and enhances current infectious disease analysis pipelines and
encourages comparisons across models and epidemics. A major contribution
of this work is the set of novel \textit{time-invariate} tools for model
and epidemic comparisons - including time-invariate prediction bands.
\pkg{EpiCompare} embraces \proglang{R}'s \textit{tidy} coding style to
make adoption of the package easier and analysis faster. This paper
provides an overview of both the tools in and intuition behind
\pkg{EpiCompare} and a thorough demonstrating of the tools through a
detailed example of a full data analysis pipeline.
}

@@

\usepackage{amsmath} \usepackage{amsthm} \usepackage{afterpage}

@@

\textcolor{violet}{I something wory about including a full paragraph of introduction before we emphasis where and why we think this tool is useful. Especially the first "where" it could be useful, as I'm not sure how it connets to the more "global" view of the epidemic research area.}

@@

\pkg{EpiCompare} also emphasizes the value of analyzing epidemics in a
\textit{time-invariant} way. Epidemics, despite by definition being a
process that evolves over time, often need to be compared in a way not
constrained to initial times or time scales to understand the processes
at play.
\textcolor{violet}{Moreover, many current comparison tools for state-space models (e.g. SIR models) highlight the proportion of individuals in each state (at a given time) in a piece-wise / marginal  fashion. This approach may reduce the amount of connections that can be seen, similar to projections of a multidimensional distribution onto a single axis at a time.}
Tools in \pkg{EpiCompare} give the user the ability to extend their
toolkit to evaluate epidemics within a time-invariant lens. The goal of
\pkg{EpiCompare} is not to supplant existing infectious disease modeling
tools and software but, rather, is a concerted effort to create standard
and fair comparisons among models developed for disease outbreaks and
outbreak data.

This paper is broken up into the following sections; section
\ref{sec:time-invariant} motivates and showcases tools of time-invariant
analysis, section \ref{sec:overview} presents an outline of how
\pkg{EpiCompare} aids a practitioner in every step of the pipeline and
section \ref{sec:tour} provides a thorough demonstrating of the tools
through a detailed example of a full data analysis pipeline.

@@

\textcolor{violet}{the vast majority of this section has been changed / needs more updates.}

\pkg{EpiCompare} delivers \textit{time-invariance} centric analysis by
(1) taking a global, not marginal view of how epidemics move through
time and (2) by treating full epidemics as filaments, not functions
(breaking away from a time-centric view). The following section aims to
highlight the strengths of \textit{time-invariance} and define the
mathematical foundations that \pkg{EpiCompare}'s tools stand upon.

Epidemics are complex objects, and hard to assess and compare to one
another due to the differences in the diseases, the location where the
outbreak occurs, how the effected population reacts, and the time
aspects (including start of the epidemic, speed of infection and more).
Time-invariant analysis attempts to make different epidemics easier to
compare by removing many time dependent aspects of an epidemic and
instead focus on the overall ``life'' of an epidemic, emphasizing the
number of lives affected.

\subsection[motivating through R0]{Motivating time-invariance through
\(R_0\)}\label{r0}

Time-invariance analysis, as it appears in \pkg{EpiCompare}, can be seen
as an attempt to escape many difficulties comparing different epidemics.
With time-invariant analysis, comparing the decades-long outbreak of HIV
in the US to a 10 day outbreak of norovirus on a cruise ship is still
possible. Time-dependent problems can arise when estimation of certain
epidemiological parameters, including the reproductive number \(R_0\),
which \citet{Gallagher2020} showed can be heavily impacted by choices of
beginning and ending time points of an epidemic.

\textcolor{violet}{
Story: 
(1) basic definition of $R_0$, (2) emphasis that $R_0$ captures a lot of information about an epidemic. Given 3 number example and discuss how one would usually interpret what the differences in $R_0$ meant (3) maybe connect to \# of $R_0$ estimates for COVID-19 (Aronson et al 2020). 
}

\(R_0\) is probably the most famous time-invariant numerical summary of
an epidemic and is associated with the Susceptible-Infectious-Recovered
(SIR) model. For a demonstration of \(R_0\) getting lost in a
time-dependent analysis we introduce the \citet{Kermack1927}'s common
SIR model. This model captures the transitions from one state to the
next as a system of ordinary differential equations, where \(N\) is the
total number of individuals, \(\beta\) is the rate of infection, and
\(\gamma\) is the rate of recovery,

\begin{align}\label{eq:sir-ode}
    S^\prime(t) &= -\frac{\beta S(t)I(t)}{N} \\
    I^\prime(t) &= \frac{\beta S(t)I(t)}{N} - \gamma I(t) \nonumber\\
    R^\prime(t) &= \gamma I(t) \nonumber.
\end{align}

From this model, \(R_0 = \beta/\gamma\), aka the ratio of the estimated
infection rate compared to the estimated recovery rate.

Interestingly, with regards to traditional epidemic \(state\)
vs.~\(time\) plots, when comparing different epidemics their \(R_0\)s
are often impossible to compare. This can be seen is a basic example,
where we have 2 different epidemics that start in a population of 1000
with 10 individuals infected, but with different parameters related to
equation \ref{eq:ODE}. Specifically suppose that
\(\beta_1, \gamma_1 = (a,b)\) and \(\beta_2, \gamma_2 = (c,d)\) and that
we observe both of these epidemics for 15 days. The epidemic
trajectories are shown in the \(state\) vs.~time plots in Figure
\ref{fig:different-scales-standard}. At a glance, we may assume that
Model 1 has a larger \(R_0\) than Model 2 because the peak of infection
occurs more quickly than in Model 2. On the other hand, we may think
Model 2 has a larger \(R_0\) because we may think the number of
infections in that model has not yet peaked at time 15. Yet both
epidemics have the same \(R_0\) of 2.

However, when we treat both epidemics a functional data through a space
that tells us the proportion of the population each each class, as is
seen in figure 3 we see a different story. The points seem to overlap
and form the same trajectory. Now it seems to be that Model 2 is
following the same trajectory as Model 1 but is not as far along in the
infection process. We can see there is something fundamentally linking
these two different epidemics, and this fundamental link turns out to be
\(R_0\).

We can see this mathematically if we let our two epidemics be presented
as \(\{(S_1(t), I_1(t), R_1(t))\}_{t\geq0}\),
\(\{(S_2(s), I_2(s), R_2(s))\}_{s \geq 0}\) respectively. As with the
example, assume both models have the same initial values
\((S(0), I(0), R(0))\), and let
\(R_0 =\frac{\beta_1}{\gamma_1} = \frac{\beta_2}{\gamma_2}\) where
\(\beta_i\) and \(\gamma_i\) are the average infection rate and recovery
rate, respectively, for SIR model \(i=1, 2\). And define \(a>0\) to be
the relative scalar such that \(\beta_2 = a \beta_1\) if and only if
\(\gamma_2 = a \gamma_1\).

The proof of Theorem \ref{thm:sir-scale} relies on a fairly recent
result from \cite{Harko2014} and is shown in detail in Proof
\ref{proof:thm}. The consequence of Theorem \ref{thm:sir-scale} is that
for two SIR models that have the same initial percent of individuals in
each state and \(R_0\) then for every point on the epidemic path of the
first SIR model is also a point on the epidemic path of the second SIR
model.

@@

\subsection[Beyond R0 and SIR]{Beyond \(R_0\) and Kerman's and
McKendrick SIR Models}\label{beyond-r0-sir}

Through \(R_0\) we saw that treating epidemics more like a function of
the time points in some higher dimensional space (nicely visualizable
with a ternary plot for the SIR models), then we could better compare
the overall structure of the epidemic and how it impacted the
population. Without knowing the generative structure underlying these
two epidemics, we propose that time-invariant tools can help compare
them in smart ways, and that these ideas can extend past the CITE
kermack1927 model and to epidemics with more states.

\textcolor{violet}{
to be continued - (1) highlight filamental view of the epidemics through their simplex space (2) describe distances that could compare different epidemics with this mindset. Also motivate uncertainty for epidemics now that we have a filamental view (3) emphasis that this idea should extend to higher dimensions (4) present the idea that one could also compare bands together.
}
